---
title: Linking calculated and perceved accessibility to suggest equity standards 
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliation: Some Institute of Technology
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    organization: Big Wig University
    addressline: 1 main street
    city: Gotham
    state: State
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  This is the abstract.

  It consists of two paragraphs.
keywords: 
  - keyword1
  - keyword2
journal: "An awesome journal"
date: "`r Sys.Date()`"
linenumbers: false
numbersections: true
bibliography: mybibfile.bib
biblio-style: elsarticle-harv # author year style for natbib - use 'elsarticle-num' or 'elsarticle-num-names' for numbered scheme
classoption: preprint, 3p, authoryear # remove authoryear is not using `elsarticle-harv`
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

```{r}
library(survey)
library(svydiags)
```


# Introduction 

Transportation equity refers to the distribution of the burdens and benefits of the transportation system among the population (Karner et al., 2020, 2024; Pereira et al., 2017; Pereira & Karner, 2021). Burdens include health-related concerns arising from exposure to air pollution, noise, and traffic; sedentary lifestyles; and the risk of injury or death from collisions. In contrast, the primary benefit is the improvement of population accessibility (Pereira & Karner, 2021), understood as the ease of reaching (spatially distributed) opportunities and destinations (5–8), such as employment, healthcare, educational services, and other places for social interaction. 

Accessibility is one of the primary outcomes of spatial development (Páez et al., 2012). It is influenced by the spatial distribution of the population and opportunities and depends on the characteristics and performance of the transportation network, as a product of both transportation systems and land use patterns (Wee & Geurs, 2011). There is growing consensus among researchers and transportation planners that the goal of transport policy should be to improve accessibility. This contrasts with traditional planning approaches that have prioritized mobility patterns that stimulate private automobile use and higher speeds as solutions to congestion and long commuting times. Mobility-centered policies often focus on quantitative metrics, such as increasing the number of trips, raising average speeds, and reducing congestion, treating mobility as an end in itself. However, this approach is problematic because people rarely travel simply for the sake of moving; instead, they travel to access opportunities, activities, or other people at their destinations. Therefore, since the main goal is to enable people to access activities, land use and transport planning practices must be redesigned to prioritize accessibility.

Insufficient or inadequate accessibility can prevent individuals from participating in location-based activities they wish to engage in for reasons beyond their control, in a process described in the literature as Transport-Related Social Exclusion (TRSE) (Casas, 2007; Lucas, 2012; Paez et al., 2009; Preston & Rajé, 2007; Luz & Portugal, 2022). 

Insufficient accessibility, when combined with other forms of social and economic disadvantage, can result in transportation poverty (Casas, 2007; Litman, 2024; Lucas, 2012, 2019; Paez et al., 2009; Preston & Rajé, 2007): the failure of transportation and land use systems to provide access to essential services or employment (Luz & Portugal, 2022). Transportation poverty increases vulnerability among individuals who already face challenges associated with age (van Dülmen et al., 2022), income (Allen & Farber, 2019), race (Awaworyi Churchill, 2020), gender (Murphy et al., 2023), disability (Upham et al., 2022), among other factors (Kim et al., 2024).

With the gentrification of Canadian downtowns and the suburbanization of low-income populations, it is estimated that over one million Canadians live in transport poverty (Allen & Farber, 2019). An equity standard, when operationalized, defines the level of fairness to be adopted and the acceptable inequality (Soukhov, Anastasia et al., 2024). However, the lack of standards for the equitable distribution of transportation benefits complicates efforts to address transportation poverty.

Additionally, there is an absence of transportation data in the form of open data products to support urban planners (Arribas-Bel et al., 2021; Behrendt & Sheller, 2021; Páez, 2021), and transportation planners receive limited formal guidance in assessing the equity implications of their projects (Miller, 2018; Palm et al., 2023; te Brömmelstroet et al., 2016). Lastly, few studies explicitly link accessibility levels with employment outcomes, such as employment opportunities declined or accepted due to transport conditions (Soukhov, Anastasia et al., 2024), or individuals' perceptions - for instance, their opinion on how easy is to reach employment opportunities and whether jobs are available within an ideal commuting time. Combined, this results in a gap in the development and implementation of accessibility standards to ensure equity in transportation planning and policy.

This study is guided by the following research questions: How can calculated job accessibility be related to perceived accessibility and employment outcomes? And could a given level of accessibility serve as a standard or goal for equitable transport planning? To answer these questions, this study has the general objective of identifying key thresholds in the relationship between calculated job accessibility and people's perceptions of employment outcomes in order to define equity standards for transportation planning.

Our research advances the current understanding of the literature on the equitable distribution of transportation benefits by proposing a method that identifies calculated accessibility thresholds based on their relationship with perceived accessibility. Identifying minimum job accessibility levels in practice would facilitate the implementation of normative criteria that can guarantee standards to prevent equity-deserving groups from being excluded from the labour market. Furthermore, accessibility thresholds can provide policymakers with an understanding of how much accessibility is needed to meet the fundamental needs of the population, increasing the potential for government adoption and enhancing the comprehensibility and relevance of standards for community members and those experiencing unequal mobility.

We also present, in a transparent manner, the moral guidelines that support the distribution of accessibility. In addition, to facilitate collaboration and further analysis and to ensure the transparency and replicability of our study, we developed this paper using literate programming, with the data analysis code accessible through our GitHub page (https://github.com/dias-bruno/NSTP), in alignment with best practices in spatial data science (Arribas-Bel et al., 2021).

# Theoretical Background 

## Calculated and perceived accessibility 

The most commonly used accessibility measures are based on the gravity model of Hansen and make use of spatial data. This model considers the number of opportunities that a person can reach to be proportional to the sum of opportunities in each destination, discounted by the travel cost to reach that destination. This type of metric is classified as a location-based measure, representing a characteristic of a given location that assumes that all people in this location have equal access to the spatially distributed opportunities. Such measures are sensitive to land use factors and to the performance of the transportation system (accessibility 2004). However, they do not take into account the individual characteristics of people.

<!-- Apart from Hanson's (1959) innovative accessibility measure, this method not consider competition in the distribution of opportunities. Also, this measure is considered an unconstrained measure, which means that the same opportunity can be allocated to different origins, making it difficult to interpret the result. Another difficulty in interpreting the result is caused by the way in which the number of job opportunities is discounted by the cost of travel. -->

The perceived accessibility, on the other hand, refers to the perception of people of how easily they can reach opportunities (Lattman et al., 2016; Pot et al., 2021), resulting from not only the interaction of land-use and transport infrastructure, but also from travel options (e.g., having a car or a transit pass), travel experiences, and individual characteristics and preferences (e.g., Lattman et al., 2016; Mehdizadeh 2025). Often, perceived accessibility deviates from place-based calculated accessibility, with some research indicating that calculated accessibility might not be able to capture the multiple ways in which people experience a lack of access to opportunities (Zhu & Lucas, 2025; Paromita, 2026). Explanations for these differences are partially attributed to differences in individuals' travel attitudes, capabilities, and preferences (J. De Vos et al. 2025 - https://www-sciencedirect-com.libaccess.lib.mcmaster.ca/science/article/pii/S0965856425001818?via%3Dihub; van der Vlugt et al., 2022). .

Even though, as demonstrated in the following subsection, calculated accessibility is still commonly used to evaluate the effect that the interaction between land-use characteristics and the transportation system has on employment outcomes. In addition, existing studies on perceived accessibility seem to have difficulties in creating standardized definitions and measurement tools, contributing to the somewhat limited knowledge on the topic (e.g., Negm et al., 2025). In this paper, as suggested by Pot et al. (2024), we define sufficient levels of accessibility based on the relationship between calculated accessibility and people's perception of employment outcomes, while also controlling for individual characteristics.

## Job accessibility and employment outcomes

Studies connecting employment outcomes to job accessibility date back to the 1960s, when John Kain (1968) proposed the spatial mismatch hypothesis while analyzing Detroit and Chicago, USA. The author hypothesized that the low employment rates among Black Americans in inner cities resulted from employment decentralization and the difficulty faced by the Black population in moving out of these areas, primarily due to racial discrimination and segregation. 

Since the work of Kain , studies linking calculated job accessibility and different employment outcomes have presented mixed results. Bania et al. (2008) found no significant effect for job access when examining its relationship with employment, earnings, hourly wage, and weekly hours worked for women leaving welfare in Ohio, USA. Delmelle et al. (2021) also found that changes in car accessibility had no significant impact on employment rates for different population income groups in the Metropolitan Area of Charlotte, North Carolina, USA, arguing that the spatial separation between lower-wage workers and their workplaces might already be relatively small. Studying Los Angeles, USA, Merlin and Hu (2017) found mixed-effects of job accessibility on employment status, with a stronger association for more highly educated groups. The authors argued that lower-educated groups may be more limited by non-spatial factors that impede their employment opportunities, such as qualifications and social skills. On the other hand, for workers with a higher level of education, the time and monetary costs of spatial separation itself may play a relatively more important role, so accessibility measures may have a greater impact on their decisions and ability to enter the labor market.

In contrast, other studies found significant effects of job accessibility on employment outcomes (Bastiaanssen et al., 2021; Duarte et al., 2023; Jin & Paulsen, 2018; Johnson et al., 2017; Matas et al., 2010). Jin and Paulsen (2017) used a difference-in-differences model for the Chicago Metropolitan Area, finding that increasing job accessibility between 2010 and 2017 was associated with reduced unemployment rates. Duarte et al. (2023) found that job accessibility had a positive effect on participation in the labour market for women, a negative effect on informal employment for women, and a positive effect on the probability of being employed for both women and men. More recently, Bastiaanssen et al. (2025) examined differential employment effects of job accessibility by public transport in combination with the bicycle for different educational groups at the national level in the Netherlands. They found that jobs for lower-educated groups are often located outside areas of high accessibility, thereby reducing job accessibility among these groups, while jobs for higher-educated groups tend to be concentrated in and around downtowns and are associated with greater accessibility.

## Equity Standards

Accessibility standards are often operationalized as a type of opportunity-based standard, defined using thresholds in access to opportunities (e.g., the number of opportunities within a 10-minute walking trip) or based on travel impedance thresholds (e.g., a given distance to a subway station). After a systematic review, Souhkov et al. identified that this type of standard is the most commonly applied in transportation literature (68%).

Yenisetty & Bahadure (2020) set an opportunity-based standard, considering a distance of more than 1,200 meters between amenities, services, and facilities by public transport to be unfair when analyzing cities in India. Tiznado-Aitken (2018) defined unfair cases as those where bus stops and metro stations are located more than 1 kilometer away. Sun & Thakuriah adopted 400 meters as the maximum acceptable walking distance to bus and tram stops, and 800 meters to ferry stations in England. In another study, Peungnumsai et al. (2020) identified opportunity standards through the ratio of public transport supply to demand, considering conditions fair when demand is greater than supply.

However, one must to consider the possible implications of relying on normative values to define thresholds. For example, when adopting a fixed distance to a certain bus stops or train stations the justification for this choice is often unclear. In addition, the use of distance-based thresholds may be problematic, since the time required to cross a certain distance can vary considerably within a city, even for the same mode of transportation.

# Materials and Methods

The methodology for obtaining standards of accessibility based on the relationship between calculated measures and people's perceptions can be synthesized in six main steps: modelling job accessibility, selecting population groups with the lowest perceived accessibility, predicting employment outcomes using calculated accessibility, setting equity standards for transportation, and defining priority geographical areas for policy interventions. We recognize accessibility as the transport benefit to be distributed across society, but we also find it necessary to define the moral principles guiding such distribution, how the distribution should occur to ensure fairness, and the requirements for considering a population group a priority group. Therefore, the following guidelines have been defined:

1. We recognize accessibility as a human capability, and a higher level of accessibility implies more choice.
2. A fair transportation and land-use system provides sufficient accessibility to all members of society. A minimum level of job accessibility is essential for individuals to meet their basic needs and, while not sufficient on its own, it is a prerequisite for promoting equality of job opportunities (sufficientarian perspective).
3. The minimum level of accessibility should be defined in terms of the most unsatisfied population group in relation to the transportation system, while considering equity perspectives and implications.
4. A fair distribution of accessibility should focus on reducing inequalities of opportunity by prioritizing disadvantaged populations, while simultaneously aiming to improve overall accessibility levels.
5. Although accessibility standards should be based on the needs of disadvantaged populations, it is also necessary to identify areas where public policy interventions should be targeted, as the needs of populations can vary significantly depending on individual constraints.

## Study Area and Data Sources

Our study area refers to the urban conurbation formed by the Greater Toronto Area and the Hamilton Area (GTHA), in Ontario, Canada. More specifically, the study area refers to the Census Divisions of the regional municipalities of Durham, Hamilton, Halton, Peel, Toronto, and York. According to the latest census (2021), over 7.2 million residents live in this area, which is spread across around 7,300 square kilometers.

Regarding the primary data sources, this study uses data from the 2021 Census of Population (CoP), from Statistics Canada, and from the National Survey of Transport Poverty (NSTP), from the Mobilizing Justice Partnership. The Census was selected due to its various socio-economic and demographic variables and due to the "Commuting to Work" questionnaire. The Census allows the identification of individual in the labour force and include variables required to calculate job accessibility, such as place of work, commuting modes, and commuting duration. The Census also provides standardized spatial datasets with statistical and administrative geographical boundaries, including the Census Tract (CTs), which are used in this study as the spatial units used for calculating accessibility measures. According to the Census, 3,868,120 individuals in the labour force reside in the study area, accessing a total of 3,043,730 job opportunities. Figure 2 shows the total labour force and the number of jobs across census tracts in the study area.

The second survey, NSTP, is a large-scale survey designed to understand transport-related social exclusion (TRSE) and transport poverty in Canada, collecting responses from more than 27,000 people across the country and providing an unprecedented look at the transportation challenges Canadians face, especially among equity-deserving groups whose experiences are often overlooked. The survey is openly available to researchers, decision-makers, and the public, and offers information on household dynamics, respondents' perceptions, and individual-level socioeconomic data. The NSTP provides respondent weights to enable the creation of statistics that are consistent and representative of the Canadian population (Morency et al., 2024). This survey was selected because it contains respondents' perceptions of accessibility and employment outcomes, as well as socioeconomic and demographic variables at the individual level, with a geographic resolution that enables the linking with calculated accessibility. The NSTP sample selected for this study includes all respondents residing in the GTHA, totalling 3,243 respondents that represent 4,890,147 individuals in the population (weighted sample).

We selected the NTSP respondents' responses to the following affirmations, which were answered using a continuum scale from 0 (completely disagree) to 100 (completely agree): “To what extent do you agree with the following statements?”: 

- It takes too long to travel to work;
- I have had to decline employment opportunities because of my transport situation;
- With the transport options available to me, I can easily reach work;
- There are too few jobs available for me in my ideal commuting time.

## Modelling accessibility

In this study, we implement the multimodal spatial availability measure to model job accessibility. This measure is an extension of the spatial availability measure to include travel by different modes. Both measures consider the competition for opportunities, which avoids a single job opportunity being equally available to more than one person. Competitive measures matter when considering employment outcomes because jobs are exclusive, and their use by one person prevents their use by another. In addition, these measures relies on a proportional allocation that balances the cost of travel and the population size, imposing a constrained calculation that allocates the opportunities to each origin ensuring that, when all jobs allocated to an origin are summed, they equal the total number of opportunities in the region. Lastly,  both measures have the advantage of resulting in an interpretable indicator, in which the output of the model refers to the number of jobs reachable in each spatial unit.

We calculated the multimodal spatial availability $V_i^m$ using Equation (\ref{eq:spatial-availability-multimodal}): 

```{=tex}
\begin{equation}
\label{eq:spatial-availability-multimodal}
V^m_{i} = \sum_{j=1}^J O_j\ F^{tm}_{ij}
\end{equation}
```

\nonident where: 

- $V^m_{i}$ is the spatial availability of opportunities at origin $i$ and travel mode $m$;
- $O_j$ is the number of opportunities at destination $j$;
- $F^{tm}_{ij}$ is a balancing factor allocating opportunities at $j$ to origin $i$ considering mode $m$. 

$F^{tm}_{ij}$ incorporates both population demand and travel impedance, defined according to Equation {eq:multimodal-balancing-factors}: 

```{=tex}
\begin{equation}
\label{eq:multimodal-balancing-factors}
F^{tm}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m=1}^M \sum_{i=1}^N F^{pm}_{i} \cdot F^{cm}_{ij}}
\end{equation}
```

The population-based allocation factor is obtained by: 

- $F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}$

\nonident where $P_{i}^m$ represents the population at origin $i$ using mode $m$. 

On its turn, the travel impedance allocation factor is given by:

- $F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}$

\nonident where: 

- $c^m_{ij}$ is the cost to travel between $i$ and $j$ for mode $m$; 
- $f^m{c^m_{ij}$ is an impedance function representing the effect of travel cost. 

Using microdata from the 21CoP, we calculated for each CT the number of people in the labour force that commute to work by walking, cycling, car-motorized modes (those who commute by car, truck, or van as a driver or passenger, and motorcycles or scooters), and public transit (those who commute by bus, subway, light rail, streetcar, commuter train, and passenger ferry). We distributed remote workers according to the share of each transportation mode in each CT. Then, we summed the number of jobs in each CT considering the place of work mentioned by each worker.

We calculated the travel times between origins and destinations for each travel mode using the R package r5r with a street network obtained from OpenStreetMap. We used the General Transit Feed Specification (GTFS) obtained from the Canadian Public Transit Network Database to retrieve the public transit schedule and coverage. Considering that our study area contains different transit agencies, we evaluated the GTFS files to identify a common date/week with public transit service. After this analysis, we set 8 a.m. (morning peak time) of December 22, 2025, as the trip departure time. We also used Digital Elevation Models (DEM) from the Canadian Digital Surface Model (CDSM), with a spatial resolution of about 20 meters, to estimate the effect of declivity on walking and cycling times. We set a maximum duration of 180 minutes (3 hours) for car and public transit trips, with a maximum walking time of 5 minutes to reach the vehicle and 15 minutes to reach the public transit stop/station; for walking and cycling trips, we set a maximum duration of 90 minutes (one and a half hours). After estimating the travel times between CTs, we assumed that all intra-CT trip durations are equal to 0.1 minute.

Commute duration was used to calibrate Probability Density Functions (PDFs) and represent travel impedance, following the approach of Soukhov and Páez [-@soukhov2024], Verma and Ukkusuri [-@sveda2023], and Dos Santos et al. [ACTIVE TRAVEL PAPER]. We applied the fitdistrplus package [@delignette2015fitdistrplus] and tested different functional forms (uniform, negative exponential, gamma, normal, and lognormal distributions) to select the best PDF for each mode of transportation and municipality, using a selection criteria the lowest Akaike Information Criterion (AIC) value [@akaike1974] (Appendix 1 displays the estimated impedance functions).

## Modelling employment outcomes

We combined the multimodal spatial availability measures with variables at the individual-level from the NTSP to model perceptions on employment outcomes. Considering each respondent weight, we transformed the respondents' affirmation to a likert scale based on their quintile position, categorizing respondent's perceptions between: "Most Negative", "Negative", "Neutral", "Positive", and  "Most Positive".

We employed weighted ordinal logistic regression models to examine the effect of between the accessibility indicators on the respondents' perceptions, using the Equation {eq:ordinal-logit}:  

```{=tex}
\begin{equation}
\label{eq:ordinal-logit}
\log(\frac{P(Y\le j)}{P(Y\ge j)}) = \theta_j - \beta_1 X_1 + \beta_2 X_2 + ... + \beta_p X_p
\end{equation}
```

\nonident where $Y$ represents the perception category of the respondent (ranging from "Most Negative" to "Most Positive"), and $X_1, X2, ..., X_p$ represent the set of explanatory variables, including multimodal spatial availability measures and individual and household characteristic obtained from the NTSP. $\theta_j$ and $\beta_p$ are unknown parameters to be estimated. The parameters $\beta_p$ are constant for all outcome categories and measure the change in the log-oggs of reporting higher perception categories associated with each explanatory variable, while $\theta_j$ represents the intercept separating the perception categories. We fitted weighted logistic models using the `survey` `R` library (Lumley 2010) to account for sampling weights and to obtain estimates representative of the study area population. 

Table X 

```{r}
respondents <- readRDS("C:/Bruno/NSTP/data-raw/output/GTAH/CT/standards-measures-per-person/easy_to_reach_ordinal.rds")

respondents$call
```


## Controlling for endogeneity



## Equity Standards 

## FGTs 

## Areas for priority intervention 

# Results and discussions 

# Limitations

# Summary and conclusions 




# References {-}

