---
title: "Exploring equity standards for transport policy: A case study of job accessibility in the Greater Toronto and Hamilton Area, Canada"
author:
  - name: Bob Security
    email: bob@example.com
    affiliation: Some Institute of Technology
    correspondingauthor: true
    footnote: 1
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
address:
  - code: Some Institute of Technology
    organization: Big Wig University
    addressline: 1 main street
    city: Gotham
    state: State
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
journal: "An awesome journal"
date: "`r Sys.Date()`"
linenumbers: false
numbersections: true
bibliography: mybibfile.bib
biblio-style: elsarticle-harv # author year style for natbib - use 'elsarticle-num' or 'elsarticle-num-names' for numbered scheme
classoption: preprint, 3p, authoryear # remove authoryear is not using `elsarticle-harv`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
header-includes:
  - \usepackage{pdflscape}
editor_options: 
  markdown: 
    wrap: 72
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  warning = FALSE,
  comment = '', 
  out.width = "1\\linewidth")

knitr::knit_hooks$set(inline = function(x) {
  if (is.numeric(x)) {
    format(x, big.mark = ",", scientific = FALSE)
  } else {
    x
  }
})

opts <- options(knitr.kable.NA = "")

rm(list = ls())
```

```{r load-packages, include=FALSE, cache=FALSE}
library(here)
library(kableExtra)
library(flextable)
library(dplyr)
```

```{r load-data, include = FALSE}
# Design 
standards_measures_diretorio <- paste0(here::here(),"/data-raw/output/GTAH/CT/standards-measures-per-person/")
```

# Supplementary material

Table \ref{tab:pcsd_summary} presents the total number of people living in each priority level for policy intervention, by equity standard type and census subdivision. Priority areas for policy intervention are determined with reference to the transit accessibility shortfall and to the Accessibility Fairness Index based on conservative, intermediate, and progressive equity standards.

```{r}
summary_pcsd <- read.csv(paste0(standards_measures_diretorio, "/summary_pcsd_priority_areas.csv")) %>%
  dplyr::select("pcsd_label", "AFI_type", "Priority_value",  "Population", "Pop_percent", "CTs","CT_percent") %>%
  mutate(Pop_percent = round(Pop_percent, 2),
         CT_percent = round(CT_percent, 2)) %>%
  kable(format = "latex",
        booktabs = TRUE, longtable = TRUE,
        caption = paste0("\\label{tab:pcsd_summary}People living in priority areas for policy-intervention by equity standard and census subdivision."),
        col.names = c("Subdivision", "Equity standard","Priority class", "Total","(%)","Total","(%)")) %>%
    add_header_above(c(" " = 3, "Population" = 2, "Census Tract" = 2)) %>%
    kable_styling(full_width = FALSE,
                latex_options=c("scale_down", "repeat_header"),
                font_size = 10) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, width = "5cm") %>%
  collapse_rows(columns = c(1,2), valign = "middle", latex_hline = "major")
```


```{r echo=FALSE, cache=FALSE, warning=FALSE}
summary_pcsd
```
