---
title: "Exploring equity standards for transport policy: A case study of calculated and perceived accessibility in the Greater Toronto and Hamilton Area, Canada"
author:
  - name: Bruno Dias dos Santos
    email: dossanb@mcmaster.ca
    affiliation: McMaster University
    correspondingauthor: true
  - name: Anastasia Soukhov 
    email: asoukhov@uwo.ca
    affiliation: Western University
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
address:
  - code: McMaster University
    organization: School of Earth, Environment & Society
    addressline: 1280 Main Street West
    city: Hamilton
    state: Ontario
    postcode: L8S 4K1
    country: Canada
  - code: Western University
    organization: Department of Geography and Environment
    addressline:  1151 Richmond Street
    postcode: N6A 3K7
    state: Ontario
    city: London
    country: Canada
abstract: |
  The lack of standards for the equitable distribution of accessibility complicates efforts to address transport poverty. This study identifies equity standards by exploring the relationship between calculated and perceived job accessibility for individuals living in the Greater Toronto and Hamilton Area, Canada. We use calculated accessibility indicators for three modes of travel (active transportation, public transit, and motorized modes) along individual-level variables to estimate ordinal models. To control for endogeneity we use spatial filters as instrumental variables. Two variables measure the perceptions of accessibility in the sample: the perception of having too few jobs available within an ideal commuting time and the perception of having to decline jobs because of transportation constraints. The results indicate that higher levels of active and transit accessibility tend to mitigate negative perceptions regarding these situations. Based on the results of these models, we identify accessibility thresholds that reduce the likelihood of negative perceptions for equity-deserving groups. We present three equity standards that we classify as conservative, intermediate, and progressive. We believe that these standards can serve as a data-driven example of how to support transportation planning in achieving more equitable outcomes, as well as for participatory processes to collectively define minimum or sufficient accessibility levels.

journal: "An awesome journal"
date: "`r Sys.Date()`"
linenumbers: false
numbersections: true
bibliography: mybibfile.bib
biblio-style: elsarticle-harv # author year style for natbib - use 'elsarticle-num' or 'elsarticle-num-names' for numbered scheme
classoption: preprint, 3p, authoryear # remove authoryear is not using `elsarticle-harv`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
header-includes:
  - \usepackage{pdflscape}
editor_options: 
  markdown: 
    wrap: 72
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  warning = FALSE,
  comment = '', 
  out.width = "1\\linewidth")

knitr::knit_hooks$set(inline = function(x) {
  if (is.numeric(x)) {
    format(x, big.mark = ",", scientific = FALSE)
  } else {
    x
  }
})

opts <- options(knitr.kable.NA = "")

rm(list = ls())
```

```{r load-packages, include=FALSE, cache=FALSE}
library(here)
library(kableExtra)
library(flextable)
library(dplyr)
```

```{r load-data, include = FALSE}
# Design 
standards_measures_diretorio <- paste0(here::here(),"/data-raw/output/GTAH/CT/standards-measures-per-person/")
```

# Supplementary material

Table \ref{tab:pcsd_summary} presents the total number of people in low-income living in each priority level for policy intervention, by equity standard type and census subdivision. Priority areas for policy intervention are determined with reference to the transit accessibility shortfall and to the Equity-Weighted Accessibility Fairness Index based on conservative, intermediate, and progressive equity standards.

```{r}
summary_pcsd <- read.csv(paste0(standards_measures_diretorio, "/summary_pcsd_priority_areas.csv")) %>%
  dplyr::select("pcsd_label", "AFI_type", "Priority_value",  "Population", "Pop_percent", "CTs","CT_percent") %>%
  mutate(Pop_percent = round(Pop_percent, 2),
         CT_percent = round(CT_percent, 2)) %>%
  kable(format = "latex",
        booktabs = TRUE, longtable = TRUE,
        caption = paste0("\\label{tab:pcsd_summary}People living in priority areas for policy-intervention by equity standard and census subdivision."),
        col.names = c("Subdivision", "Equity standard","Priority class", "Total","(%)","Total","(%)")) %>%
    add_header_above(c(" " = 3, "Low-income population" = 2, "Census Tract" = 2)) %>%
    kable_styling(full_width = FALSE,
                latex_options=c("scale_down", "repeat_header"),
                font_size = 10) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, width = "5cm") %>%
  collapse_rows(columns = c(1,2), valign = "middle", latex_hline = "major")
```


```{r echo=FALSE, cache=FALSE, warning=FALSE}
summary_pcsd
```
