---
title: "modelling_variables"
output: html_document
---

# Regressions: First stage

We will start the first stage of regression by creating a formula to regress accessibility based on spatial filters and other variables. 

```{r}
design_respondents <- svydesign(
  ids = ~1,  # No clustering
  weights = ~fexp_final,
  data = respondents)

stage1_formula <- as.formula(
  Accessibility ~ Sf_dist_SA_pp_Active +  
    Sf_dist_SA_pp_Transit + 
    `Sf_dist_SA_pp_Car_motor` + 
 
                                ## Transport Variables  
                                CarNumber +  
                                BikeNumber +
                                Transport_3mode + 
                                Match_mode + 
                                
                                ## Socio and demographic variables
                                Indigenous_or_racialized +  
                                Temporary_or_refugee +  
                                LongTermConditions + 
                                Woman_or_genderdiverse +   
                                Age +  
                                hh_size  + 
                                hh_income4 + 
    Pop_density + 
    BuiltV_median + SMOD_type
  )
```

Creating the first stage for transit accessibility: 

```{r}
stage1_transit_SA_pp <- svyglm(update(stage1_formula, SA_pp_Transit ~ .), design = design_respondents)

summary(stage1_transit_SA_pp, diagnostics = TRUE)
```

Creating the first stage for Car-motorized accessibility: 

```{r}
stage1_car_SA_pp <- svyglm(update(stage1_formula, `SA_pp_Car_motor` ~ .), design = design_respondents)

summary(stage1_car_SA_pp, diagnostics = TRUE)
```

Creating the first stage for Active accessibility: 

```{r}
stage1_active_SA_pp <- svyglm(update(stage1_formula, `SA_pp_Active` ~ .), design = design_respondents)

summary(stage1_active_SA_pp)
```

Adding the residuals to the data frame: 
```{r}
respondents <- respondents %>%
  mutate(SA_pp_Transit_IV = predict(stage1_transit_SA_pp, newdata = ., type = "response"),
    SA_pp_Car_IV = predict(stage1_car_SA_pp, newdata = ., type = "response"),
    SA_pp_Active_IV = predict(stage1_active_SA_pp, newdata = ., type = "response"),
    SA_pp_Transit_Res_IV = SA_pp_Transit - SA_pp_Transit_IV,  
    SA_pp_Car_Res_IV = `SA_pp_Car_motor` - SA_pp_Car_IV,  
    SA_pp_Active_Res_IV = SA_pp_Active - SA_pp_Active_IV)
```

# Ordinal 

```{r}
ordinal_SA_pp_3mode_formula <- as.formula(Outcome ~ 
                                # Spatial Availability variables
                                SA_pp_Active  +
                                SA_pp_Transit  +
                                SA_pp_Car_motor +

                                # Transport variables
                                 
                                CarNumber +  
                                BikeNumber + 
                                #Transport_3mode + 
                                Match_mode + 
                                #TransitPass + 

                                # Socio and demographic variables
                                Indigenous_or_racialized + 
                                Temporary_or_refugee +  
                                LongTermConditions + 
                                Woman_or_genderdiverse +   
                                Age +  
                                #Child_under_10 + 
                                hh_size + 
                                hh_income4 + 
                                
                                # Environmental 
                                Pop_density +
                                 BuiltV_median + SMOD_type + 
                                
                                # # Causal-effect 
                                SA_pp_Active_Res_IV +
                                SA_pp_Transit_Res_IV +
                                SA_pp_Car_Res_IV)
```

```{r}
respondents_filtered <- respondents %>%
dplyr::select(SA_pp_Active, SA_pp_Transit, SA_pp_Car_motor, Transit_user_bin, Active_user_bin, pc_class, hh_income4, Transport_3mode, BuiltV_median, SMOD_type, SMOD_majority, 
                 CarNumber, BikeNumber, Match_mode, Indigenous_or_racialized,
                Temporary_or_refugee, LongTermConditions, Woman_or_genderdiverse,
                Age, hh_size, hh_income, Pop_density, SA_pp_Active_Res_IV,
                SA_pp_Transit_Res_IV, SA_pp_Car_Res_IV, Transit_user,
                FewJobs, FewJobs_Likert, 
                TravelTooLong, TravelTooLong_Likert,
                EasyToGo, EasyToGo_Likert, 
                DeclineEmployment, DeclineEmployment_Likert,
                fexp_final, education_level)
```

## Few jobs: Likert 

```{r}
complete_data_fj <- respondents_filtered %>% 
  filter(!is.na(FewJobs))  %>%
  mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE), 
         FewJobs_Likert = 
           cut(FewJobs,
               #breaks = Hmisc::wtd.quantile(FewJobs, weights = fexp_final, probs = c(0, 0.2, 0.4, 0.6, 0.8, 1)), 
               breaks = c(0,20,40,60,80,100), 
               labels = c("Most Positive", "Positive", "Neutral", "Negative", "Most Negative"),
               include.lowest = TRUE), 
         FewJobs_Likert = factor(FewJobs_Likert, 
                                              levels = c("Most Negative", "Negative", "Neutral","Positive", "Most Positive"),
             ordered = TRUE))

# complete_data_fj <- FewJobs_3mode_lm_SA_pp_clean_design$variables %>%
#   mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE))

vars_to_scale <- c("SA_pp_Active", "SA_pp_Transit", "SA_pp_Car_motor",
  "Pop_density", "BuiltV_median", 
  "SA_pp_Active_Res_IV", "SA_pp_Transit_Res_IV", "SA_pp_Car_Res_IV")

# vars_to_scale <- c("SA_pp_Active", "SA_pp_Transit", "SA_pp_Car_motor",
#   "Pop_density")

complete_data_scaled_fj <- complete_data_fj %>%
  mutate(across(all_of(vars_to_scale), 
                function(x) {
                  scaled <- scale(x)
                  return(as.numeric(scaled))
                }))

scale_centers_fj <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_fj[[var]]), "scaled:center")
})

scale_scales_fj <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_fj[[var]]), "scaled:scale")
})
```

```{r}
design_ordinal_norm_fj <- svydesign(
  ids = ~1,
  weights = ~fexp_normalized,
  data = complete_data_scaled_fj)
```


```{r}
FewJobs_formula_3mode_ordinal_SA <- formula(update(ordinal_SA_pp_3mode_formula, FewJobs_Likert ~ .))

FewJobs_ordinal_3modes <- svyolr(FewJobs_formula_3mode_ordinal_SA,
                                   design = design_ordinal_norm_fj)

summary(FewJobs_ordinal_3modes)
```


```{r}
car::vif(FewJobs_ordinal_3modes)
```

```{r}
survey::regTermTest(FewJobs_ordinal_3modes, "SA_pp_Transit", method = "LRT")
```

## Travel Too Long: Likert 

```{r}
complete_data_ttg <- respondents_filtered %>% 
  filter(!is.na(TravelTooLong)) %>%
  mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE))

# complete_data_ttg <- TravelTooLong_3mode_lm_SA_pp_clean_design$variables %>% 
#   mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE))

complete_data_scaled_ttg <- complete_data_ttg %>%
  mutate(across(all_of(vars_to_scale), 
                function(x) {
                  scaled <- scale(x)
                  return(as.numeric(scaled))
                }))

scale_centers_ttg <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_ttg[[var]]), "scaled:center")
})

scale_scales_ttg <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_ttg[[var]]), "scaled:scale")
})
```

```{r}
design_ordinal_norm_ttg <- svydesign(
  ids = ~1,
  weights = ~fexp_normalized,
  data = complete_data_scaled_ttg)
```

```{r}
TravelTooLong_formula_3mode_ordinal_SA <- formula(update(ordinal_SA_pp_3mode_formula, TravelTooLong_Likert ~ .))

TravelTooLong_ordinal_3modes <- svyolr(TravelTooLong_formula_3mode_ordinal_SA,
                                   design = design_ordinal_norm_ttg)

summary(TravelTooLong_ordinal_3modes)
```


## Easy to Reach: Likert 

```{r}
complete_data_etg <- respondents_filtered %>% 
  filter(!is.na(EasyToGo)) %>%
  mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE))

# complete_data_etg <- EasyToGo_3mode_lm_SA_pp_clean_design$variables %>% 
#   mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE))

complete_data_scaled_etg <- complete_data_etg %>%
  mutate(across(all_of(vars_to_scale), 
                function(x) {
                  scaled <- scale(x)
                  return(as.numeric(scaled))
                }))

scale_centers_etg <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_etg[[var]]), "scaled:center")
})

scale_scales_etg <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_etg[[var]]), "scaled:scale")
})
```

```{r}
design_ordinal_norm_etg <- svydesign(
  ids = ~1,
  weights = ~fexp_normalized,
  data = complete_data_scaled_etg)
```

```{r}
EasyToGo_formula_3mode_ordinal_SA <- formula(update(ordinal_SA_pp_3mode_formula, EasyToGo_Likert ~ .))

EasyToGo_ordinal_3modes <- svyolr(EasyToGo_formula_3mode_ordinal_SA,
                                   design = design_ordinal_norm_etg)

summary(EasyToGo_ordinal_3modes)
```

```{r}
survey::regTermTest(EasyToGo_ordinal_3modes, "SA_pp_Transit", method = "LRT")
```

## Declined Employment: Likert 

```{r}
complete_data_de <- respondents_filtered %>% 
  filter(!is.na(DeclineEmployment)) %>%
  mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE), 
         DeclineEmployment_Likert = 
           cut(DeclineEmployment,
               breaks = Hmisc::wtd.quantile(DeclineEmployment, weights = fexp_final, probs = c(0, 0.2, 0.4, 0.6, 0.8, 1)), 
               #breaks = c(0,20,40,60,80,100), 
               labels = c("Most Positive", "Positive", "Neutral", "Negative", "Most Negative"),
               include.lowest = TRUE), 
         DeclineEmployment_Likert = factor(DeclineEmployment_Likert, 
                                              levels = c("Most Negative", "Negative", "Neutral","Positive", "Most Positive"),
             ordered = TRUE))

# complete_data_de <- DeclineEmployment_3mode_lm_SA_pp_clean_design$variables %>% 
#   mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE))

complete_data_scaled_de <- complete_data_de %>%
  mutate(across(all_of(vars_to_scale), 
                function(x) {
                  scaled <- scale(x)
                  return(as.numeric(scaled))
                }))

scale_centers_de <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_de[[var]]), "scaled:center")
})

scale_scales_de <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_de[[var]]), "scaled:scale")
})
```

```{r}
design_ordinal_norm_de <- svydesign(
  ids = ~1,
  weights = ~fexp_normalized,
  data = complete_data_scaled_de)
```

```{r}
DeclineEmployment_formula_3mode_ordinal_SA <- formula(update(ordinal_SA_pp_3mode_formula, DeclineEmployment_Likert ~ .))

DeclineEmployment_ordinal_3modes <- svyolr(DeclineEmployment_formula_3mode_ordinal_SA,
                                   design = design_ordinal_norm_de)

summary(DeclineEmployment_ordinal_3modes)
```
