---
title: "modelling_variables"
output: html_document
---

# Regressions: First stage

We will start the first stage of regression by creating a formula to regress accessibility based on spatial filters and other variables. 

```{r}
design_respondents <- svydesign(
  ids = ~1,  # No clustering
  weights = ~fexp_final,
  data = respondents)
```

## Transit IV

Creating the first stage for transit accessibility: 

```{r}
stage1_transit_SA_pp <- svyglm(SA_pp_Transit ~ SA_pp_Active   +
SA_pp_Active:Active_user +
Sf_dist_SA_pp_Transit   +
Sf_dist_SA_pp_Transit:Transit_user +
SA_pp_Car_motor +
SA_pp_Car_motor:CarNumber + 
## Transport Variables
BikeNumber +
Match_mode +
TransitPass +
TransitAvailable +
## Socio and demographic variables
Indigenous_or_racialized +
Temporary_or_refugee +
LongTermConditions +
Woman_or_genderdiverse +
Age +
hh_size  +
hh_income4 +
Pop_density +
pc_class, design = design_respondents)


summary(stage1_transit_SA_pp, diagnostics = TRUE)
```

```{r}
wald_transit <- regTermTest(stage1_transit_SA_pp, ~ Sf_dist_SA_pp_Transit)
wald_transit
```

## Car IV

Creating the first stage for Car-motorized accessibility: 

```{r}
stage1_car_SA_pp <- svyglm(`SA_pp_Car_motor` ~ SA_pp_Active   +
SA_pp_Active:Active_user +
SA_pp_Transit   +
SA_pp_Transit:Transit_user +
Sf_dist_SA_pp_Car_motor +
Sf_dist_SA_pp_Car_motor:CarNumber + 
## Transport Variables
BikeNumber +
Match_mode +
TransitPass +
TransitAvailable +
## Socio and demographic variables
Indigenous_or_racialized +
Temporary_or_refugee +
LongTermConditions +
Woman_or_genderdiverse +
Age +
hh_size  +
hh_income4 +
Pop_density +
pc_class, design = design_respondents)

summary(stage1_car_SA_pp, diagnostics = TRUE)
```
```{r}
wald_car <- regTermTest(stage1_car_SA_pp, ~ Sf_dist_SA_pp_Car_motor)
wald_car
```

## Active IV

Creating the first stage for Active accessibility: 

```{r}
stage1_active_SA_pp <- svyglm(`SA_pp_Active` ~ Sf_dist_SA_pp_Active   +
Sf_dist_SA_pp_Active:Active_user +
SA_pp_Transit   +
SA_pp_Transit:Transit_user +
SA_pp_Car_motor +
SA_pp_Car_motor:CarNumber +
## Transport Variables
BikeNumber +
Match_mode +
TransitPass +
TransitAvailable +
## Socio and demographic variables
Indigenous_or_racialized +
Temporary_or_refugee +
LongTermConditions +
Woman_or_genderdiverse +
Age +
hh_size  +
hh_income4 +
Pop_density +
pc_class, design = design_respondents)

summary(stage1_active_SA_pp)
```
```{r}
wald_active <- regTermTest(stage1_active_SA_pp, ~ Sf_dist_SA_pp_Active)
wald_active
```

Adding the residuals to the data frame: 
```{r}
respondents <- respondents %>%
  mutate(SA_pp_Transit_IV = predict(stage1_transit_SA_pp, newdata = ., type = "response"),
    SA_pp_Car_IV = predict(stage1_car_SA_pp, newdata = ., type = "response"),
    SA_pp_Active_IV = predict(stage1_active_SA_pp, newdata = ., type = "response"),
    SA_pp_Transit_Res_IV = SA_pp_Transit - SA_pp_Transit_IV,  
    SA_pp_Car_Res_IV = `SA_pp_Car_motor` - SA_pp_Car_IV,  
    SA_pp_Active_Res_IV = SA_pp_Active - SA_pp_Active_IV)
```


# Ordinal 

```{r}
ordinal_SA_pp_3mode_formula <- as.formula(Outcome ~ 
                                # Spatial Availability variables
SA_pp_Active   +
SA_pp_Active:Active_user +
SA_pp_Transit   +
SA_pp_Transit:Transit_user +
SA_pp_Car_motor +
SA_pp_Car_motor:CarNumber +
# Transport variables
BikeNumber +
Match_mode +
TransitPass +
TransitAvailable +
# Socio and demographic variables
Indigenous_or_racialized +
Temporary_or_refugee +
LongTermConditions +
Woman_or_genderdiverse +
Age +
hh_size +
hh_income4 +
# Environmental
Pop_density +
pc_class +
# # Causal-effect
SA_pp_Active_Res_IV +
SA_pp_Transit_Res_IV +
SA_pp_Car_Res_IV)
```

```{r}
respondents_filtered <- respondents %>%
dplyr::select(SA_pp_Active, SA_pp_Transit, SA_pp_Car_motor, 
              Sf_dist_SA_pp_Active, Sf_dist_SA_pp_Transit, Sf_dist_SA_pp_Car_motor, 
              SA_Active, SA_Transit, SA_Car_motor, Transit_user_bin, Active_user_bin, 
               Transit_user, Active_user, TransitPass,  Car_user_bin, pc_class, hh_income4, Transport_3mode, BuiltV_median, SMOD_type, SMOD_majority,
CarNumber, BikeNumber, Match_mode, Indigenous_or_racialized,
Temporary_or_refugee, LongTermConditions, Woman_or_genderdiverse,
Age, hh_size, hh_income, Pop_density, SA_pp_Active_Res_IV,
SA_pp_Transit_Res_IV, SA_pp_Car_Res_IV, Transit_user,
FewJobs, FewJobs_Likert,
TravelTooLong, TravelTooLong_Likert,
EasyToGo, EasyToGo_Likert,
DeclineEmployment, DeclineEmployment_Likert,
fexp_final, education_level, TransitAvailable, PCD, employment_status)
```

## Few jobs: Likert 

```{r}
complete_data_fj <- respondents_filtered %>% 
  filter(!is.na(FewJobs))  %>%
  mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE), 
         SA_pp_Active_unscaled = SA_pp_Active,
         SA_pp_Car_motor_unscaled = SA_pp_Car_motor,
         SA_pp_Transit_unscaled = SA_pp_Transit)

vars_to_scale <- c("SA_pp_Active", "SA_pp_Transit", "SA_pp_Car_motor",
  "Pop_density", 
  "SA_pp_Active_Res_IV", "SA_pp_Transit_Res_IV", "SA_pp_Car_Res_IV")

# vars_to_scale <- c("SA_pp_Active", "SA_pp_Transit", "SA_pp_Car_motor",
#   "Pop_density")

complete_data_scaled_fj <- complete_data_fj %>%
  mutate(across(all_of(vars_to_scale), 
                function(x) {
                  scaled <- scale(x)
                  return(as.numeric(scaled))
                }))

scale_centers_fj <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_fj[[var]]), "scaled:center")
})

scale_scales_fj <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_fj[[var]]), "scaled:scale")
})
```

```{r}
design_ordinal_norm_fj <- svydesign(
  ids = ~1,
  weights = ~fexp_normalized,
  data = complete_data_scaled_fj)
```


```{r}
FewJobs_formula_3mode_ordinal_SA <- formula(update(ordinal_SA_pp_3mode_formula, FewJobs_Likert ~ .))

FewJobs_ordinal_3modes <- svyolr(FewJobs_formula_3mode_ordinal_SA,
                                   design = design_ordinal_norm_fj)

summary(FewJobs_ordinal_3modes)
```


```{r}
car::vif(FewJobs_ordinal_3modes)
```

```{r}
survey::regTermTest(FewJobs_ordinal_3modes, "SA_pp_Transit")
```

```{r}
active_cor <- svycor(~ Sf_dist_SA_pp_Active + FewJobs, design = design_ordinal_norm_fj)$cors[2]
transit_cor <-svycor(~ Sf_dist_SA_pp_Transit + FewJobs, design = design_ordinal_norm_fj)$cors[2]
car_cor <-svycor(~ Sf_dist_SA_pp_Car_motor + FewJobs, design = design_ordinal_norm_fj)$cors[2]

filters <- c("Active", "Transit", "Car")
values <- c(active_cor, transit_cor, car_cor)
fewjobscorr <- data.frame(Filter = filters, "FewJobs" = values)
fewjobscorr
```

## Travel Too Long: Likert 

```{r}
complete_data_ttg <- respondents_filtered %>% 
  filter(!is.na(TravelTooLong), employment_status == 1) %>%
  mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE), 
         SA_pp_Active_unscaled = SA_pp_Active,
         SA_pp_Car_motor_unscaled = SA_pp_Car_motor,
         SA_pp_Transit_unscaled = SA_pp_Transit)

# complete_data_ttg <- TravelTooLong_3mode_lm_SA_pp_clean_design$variables %>% 
#   mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE))

complete_data_scaled_ttg <- complete_data_ttg %>%
  mutate(across(all_of(vars_to_scale), 
                function(x) {
                  scaled <- scale(x)
                  return(as.numeric(scaled))
                }))

scale_centers_ttg <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_ttg[[var]]), "scaled:center")
})

scale_scales_ttg <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_ttg[[var]]), "scaled:scale")
})
```

```{r}
design_ordinal_norm_ttg <- svydesign(
  ids = ~1,
  weights = ~fexp_normalized,
  data = complete_data_scaled_ttg)
```

```{r}
TravelTooLong_formula_3mode_ordinal_SA <- formula(update(ordinal_SA_pp_3mode_formula, TravelTooLong_Likert ~ .))

TravelTooLong_ordinal_3modes <- svyolr(TravelTooLong_formula_3mode_ordinal_SA,
                                   design = design_ordinal_norm_ttg)

summary(TravelTooLong_ordinal_3modes)
```


```{r}
active_cor <- svycor(~ Sf_dist_SA_pp_Active + TravelTooLong, design = design_ordinal_norm_ttg)$cors[2]
transit_cor <- svycor(~ Sf_dist_SA_pp_Transit + TravelTooLong, design = design_ordinal_norm_ttg)$cors[2]
car_cor <- svycor(~ Sf_dist_SA_pp_Car_motor + TravelTooLong, design = design_ordinal_norm_ttg)$cors[2]

filters <- c("Active", "Transit", "Car")
values <- c(active_cor, transit_cor, car_cor)
traveltoolongcorr <- data.frame(Filter = filters, "TravelTooLong" = values)

traveltoolongcorr
```

## Easy to Reach: Likert 

```{r}
complete_data_etg <- respondents_filtered %>% 
  filter(!is.na(EasyToGo), employment_status == 1) %>%
  mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE), 
         SA_pp_Active_unscaled = SA_pp_Active,
         SA_pp_Car_motor_unscaled = SA_pp_Car_motor,
         SA_pp_Transit_unscaled = SA_pp_Transit)

# complete_data_etg <- EasyToGo_3mode_lm_SA_pp_clean_design$variables %>% 
#   mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE))

complete_data_scaled_etg <- complete_data_etg %>%
  mutate(across(all_of(vars_to_scale), 
                function(x) {
                  scaled <- scale(x)
                  return(as.numeric(scaled))
                }))

scale_centers_etg <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_etg[[var]]), "scaled:center")
})

scale_scales_etg <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_etg[[var]]), "scaled:scale")
})
```

```{r}
design_ordinal_norm_etg <- svydesign(
  ids = ~1,
  weights = ~fexp_normalized,
  data = complete_data_scaled_etg)
```

```{r}
EasyToGo_formula_3mode_ordinal_SA <- formula(update(ordinal_SA_pp_3mode_formula, EasyToGo_Likert ~ .))

EasyToGo_ordinal_3modes <- svyolr(EasyToGo_formula_3mode_ordinal_SA,
                                   design = design_ordinal_norm_etg)

summary(EasyToGo_ordinal_3modes)
```

```{r}
survey::regTermTest(EasyToGo_ordinal_3modes, "SA_pp_Transit")
```

```{r}
active_cor <- svycor(~ Sf_dist_SA_pp_Active + EasyToGo, design = design_ordinal_norm_etg)$cors[2]
transit_cor <- svycor(~ Sf_dist_SA_pp_Transit + EasyToGo, design = design_ordinal_norm_etg)$cors[2]
car_cor <- svycor(~ Sf_dist_SA_pp_Car_motor + EasyToGo, design = design_ordinal_norm_etg)$cors[2]

filters <- c("Active", "Transit", "Car")
values <- c(active_cor, transit_cor, car_cor)
easytogocorr <- data.frame(Filter = filters, "EasyToGo" = values)
easytogocorr
```


## Declined Employment: Likert 

```{r}
complete_data_de <- respondents_filtered %>% 
  filter(!is.na(DeclineEmployment)) %>%
  mutate(fexp_normalized = fexp_final / mean(fexp_final, na.rm = TRUE), 
         SA_pp_Active_unscaled = SA_pp_Active,
         SA_pp_Car_motor_unscaled = SA_pp_Car_motor,
         SA_pp_Transit_unscaled = SA_pp_Transit)


complete_data_scaled_de <- complete_data_de %>%
  mutate(across(all_of(vars_to_scale), 
                function(x) {
                  scaled <- scale(x)
                  return(as.numeric(scaled))
                }))

scale_centers_de <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_de[[var]]), "scaled:center")
})

scale_scales_de <- sapply(vars_to_scale, function(var) {
  attr(scale(complete_data_de[[var]]), "scaled:scale")
})
```

```{r}
design_ordinal_norm_de <- svydesign(
  ids = ~1,
  weights = ~fexp_normalized,
  data = complete_data_scaled_de)
```

```{r}
DeclineEmployment_formula_3mode_ordinal_SA <- formula(update(ordinal_SA_pp_3mode_formula, DeclineEmployment_Likert ~ .))

DeclineEmployment_ordinal_3modes <- svyolr(DeclineEmployment_formula_3mode_ordinal_SA,
                                   design = design_ordinal_norm_de)

summary(DeclineEmployment_ordinal_3modes)
```
```{r}
active_cor <- svycor(~ Sf_dist_SA_pp_Active + DeclineEmployment, design = design_ordinal_norm_de)$cors[2]
transit_cor <- svycor(~ Sf_dist_SA_pp_Transit + DeclineEmployment, design = design_ordinal_norm_de)$cors[2]
car_cor <- svycor(~ Sf_dist_SA_pp_Car_motor + DeclineEmployment, design = design_ordinal_norm_de)$cors[2]


filters <- c("Active", "Transit", "Car")
values <- c(active_cor, transit_cor, car_cor)
declineemploymentcorr <- data.frame(Filter = filters, "DeclineEmployment" = values)

declineemploymentcorr
```


```{r}
# de_ame_active_user <- svyAME(DeclineEmployment_ordinal_3modes,
#               varname = "SA_pp_Active",
#               diffchange = "sd",
#               byvar = "Active_user",
#               seed = 51)
# 
# plot(de_ame_active_user) +
#   theme_minimal()
```

```{r}
# de_ame_transit_user <- svyAME(DeclineEmployment_ordinal_3modes,
#               varname = "SA_pp_Transit",
#               diffchange = "sd",
#               byvar = "Transit_user",
#               seed = 51)
# 
# plot(de_ame_transit_user) +
#   theme_minimal()
```
